---
title: "Bike Crashes NC"
output: html_document
date: '2022-06-17'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Project brief
You have been working with the local government to help direct policy decisions. They are concerned about an increase in traffic accidents and think they may be able to make changes to reduce the severity of accidents that do occur. But they don’t know what the most important features are for defining the difference between severity of an accident. They would like you to investigate the data and see if it would be possible to find the factors that are most important. Based on these factors, what changes should be made to reduce the severity of accidents that do occur?


```{r include = F}
library(tidyverse)
library(readr)
library(stringr)
library(plotly)
library(ranger)
```


```{r echo=FALSE, message= F}
bikes <- read_csv("https://github.com/datacamp/careerhub-data/raw/master/Pedestrian%20and%20Bike%20Crashes/pedestrian_bike_crash.csv", 
                  na = c('Missing', 'Unknown'))%>%
  rename_all(tolower) #lower  eeverything for the constisntency
```



```{r echo=FALSE}
bikes <- bikes%>%
  select(-c(x, y, objectid_1, bikeage, bikerace, bikesex,
            crashid, drvrage, drvralcdrg, drvrrace, drvrsex,
            numbicsain, numbicscin, numbicscin, weather, numbicskil, numbicsnoi, numbicstot,
            longitude, latitude, objectid, crashid, numbicsbin,
            numbicsuin, drvragegrp, drvralcflg, county, bikeinjury, drvrinjury,crashsevr,
            drvrvehtyp, rddefects, bikealcdrg))

# The dataset has a lot of columns. Some columns are not interesting for a particular question, some columns have too many NA's. Other columns like alcohol were removed because the number of drivers who test positive for alcohol is almost non-existent. 
```



```{r include = F}
library(visdat)
#visualize missing data 
vis_miss(bikes)

# Since missing data is under 2% of all the data it's safe to just remove these rows for now 

bikes <- na.omit(bikes)
#Looks like dirung data entry somebody changed number to month for same age categories 
bikes <- bikes %>%
  mutate(bikeagegrp = str_replace_all(bikeagegrp, c(Oct = "10", Nov = "11")))%>%
          rename(ambulance = ambulancer)
```


#### Basic EDA 
I decided to go with the 'Ambulance' column as an indicator of the severity of an accident. There are some other columns, but they contain categories like 'suspected" or 'possible injury'. Here I assume that if there is an ambulance, the case is serious enough to investigate the reasons. 

```{r echo =F}
#Yearly dinamics  
ggplotly(bikes%>%
  group_by(crashyear, ambulance)%>%
  count()%>%
  ggplot(aes(x = crashyear, y = n, fill = ambulance)) + geom_col() + ylim(0, 800) +
  scale_x_continuous(breaks = c(2007:2019)) + labs(x = 'Year',
                                                   y = 'Number of crashes',
                                                   title = 'Bike crashes every year') + 
  scale_fill_manual(values=c("#31a7eb", "#F6685F")) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                                                            panel.background = element_blank(), axis.line = element_line(colour = "black")))
  
```

From looking at the number of crashes per year I can't say that we see an increase in traffic accidents. 


```{r echo = F}
#Crashes per month
bikes$crashmonth <- factor(bikes$crashmonth, levels = c("January", "February",
                                                        "March", "April", "May",
                                                        "June", 'July',
                                                        'August', 'September',
                                                        "October", "November",
                                                        "December"))
bikes%>%
  count(crashmonth)%>%
  plot_ly(x = ~crashmonth, y = ~n, color = I("#31a7eb"))%>%
  add_bars()%>%
  layout(xaxis = list(title = 'Month'),
         yaxis = list(title = 'Number of Crashes'))
```

Monthly trend resembles the average yearly temperature for North Carolina. This makes sense as warmer temperature means more people using bikes.


```{r echo = F}
#Hourly difference 
ggplotly(bikes%>%
  group_by(crashhour, ambulance)%>%
  count()%>%
  ggplot(aes(x = crashhour, y = n, fill = ambulance)) + geom_col() + 
  scale_fill_manual(values=c("#31a7eb", "#F6685F")) + labs(x = 'Hour',
                                                           y = 'Number of Crashes', 
                                                           title = 'Number of Crashes for each hour') + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                                                            panel.background = element_blank(), axis.line = element_line(colour = "black")))
```

Here we can observe not only the rise in bike crashes during the day but also the change in proportions of how many people need an ambulance. 

```{r echo = F}
#Age groups 
ggplotly(bikes%>%
  group_by(bikeagegrp, ambulance)%>%
  count()%>%
  ggplot(aes(x = bikeagegrp, y = n, fill = ambulance)) + geom_col() + ylim(0, 2000) + 
  scale_fill_manual(values=c("#31a7eb", "#F6685F")) + labs(x = 'Age Group',
                                                           y = 'Number of Crashes', 
                                                           title = 'Number of crashes per Age Group')  + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black")), tooltip = 'n')
```

Looks like in the 25-29 group people are less likely to get into accidents and then chances get higher as one gets older. 

```{r echo = F}
#Directions 
ggplotly(bikes%>%
  group_by(bikedir, ambulance)%>%
  count()%>%
  ggplot(aes(x = reorder(bikedir, n), y = n, fill = ambulance)) + geom_col() + 
  labs(x = 'Direction', y = 'Number of Crashes', title = "Biker's Riding  Direction") + 
  scale_fill_manual(values=c("#31a7eb", "#F6685F")) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                                                            panel.background = element_blank(), axis.line = element_line(colour = "black")), tooltip = 'n')
```

Most crashes happen while bikers ride along the traffic. This could mean a poorly developed biking paths system which is not surprising when we talk about US cities.


```{r echo = F}
#Positions
ggplotly(bikes%>%
  filter(bikepos %in% c(bikes %>%
                          count(bikepos)%>%
                          filter(n >100)%>%
                          pull(bikepos)),
         bikedir %in% c(bikes %>%
                          count(bikedir)%>%
                          arrange(desc(n))%>%
                          head(2)%>%
                          pull(bikedir)))%>%
  group_by(bikepos, bikedir)%>%
  count()%>%
  ggplot(aes(x = reorder(bikepos,n), y = n, fill = bikedir)) + geom_col() + ylim(0, 8000) + 
  labs(x = "Biker's Position", y = 'Number of Crashes', title = 'Where Crashes Happen',
       subtitle = "Number of crashes grouped by biker's position and direction") + coord_flip() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                                                            panel.background = element_blank(), axis.line = element_line(colour = "black")), tooltip = 'n')
```

Here we can see that indeed most crashes happen while bikers ride along the traffic and not using bike paths. Quick google search showed that even bike paths that do exist are very questionable quality. Probably digging into this could reduce accidents even more. 

```{r echo = F}
#Where most crushes occure 
ggplotly(bikes%>%
  group_by(city, ambulance)%>%
  count()%>%
  arrange(desc(n))%>%
  head(20)%>%
  ggplot(aes(x = reorder(city,n), y = n, fill = ambulance)) + geom_col() + labs(x = 'Number of Crashes',
                                                              y = 'City',
                                                              title = 'Cities with most crashes') + 
  scale_fill_manual(values=c("#31a7eb", "#F6685F")) + 
  coord_flip() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                                                            panel.background = element_blank(), axis.line = element_line(colour = "black")), tooltip = 'n')
```

Here we can see that more populous cities have more crashes which can be explained by the amount of people using bikes there. But most crashes actually happen in rural areas that lack bike infrastructure. This might be another indicator that fixing biking infrastructure can actually increase the situation. 

```{r include=FALSE}
bikes <- bikes%>%
  filter(rdconditio %in% c("Dry", 'Wet'))
  #most of the conditions are super rare I'll keep only Dry/Wet
#Road features
bikes <- bikes%>%
  filter(rdfeature %in% c(bikes%>%
  count(rdfeature)%>%
  filter(n > 40)%>%
  pull(rdfeature)))
#Crashtype 
bikes <- bikes%>%
  filter(crashtype %in% c(bikes%>%
  count(crashtype)%>%
  filter(n > 50)%>%
  pull(crashtype)))
#Lightcondition
bikes <- bikes%>%
  filter(lightcond %in% c(bikes%>%
  count(lightcond)%>%
  filter(n> 50)%>%
    pull(lightcond)))
#deffects 
bikes <- bikes%>%
  filter(rdsurface %in% c(bikes%>%
  count(rdsurface)%>%
  filter(n > 50)%>%
  pull(rdsurface)))
bikes <- bikes%>%
filter(rdcharacte %in% c(bikes%>%
  count(rdcharacte)%>%
  filter(n > 50)%>%
  pull(rdcharacte)))
#Road class
bikes <- bikes%>%
  filter(rdclass %in% c(bikes%>%
                             count(rdclass)%>%
                             filter(n > 50)%>%
                             pull(rdclass)))
#Traffic control
bikes <- bikes%>%
  filter(traffcntrl %in% c(bikes%>%
                             count(traffcntrl)%>%
                             filter( n > 50)%>%
                             pull(traffcntrl)))
#Remove cities that are very rare 
bikes <- bikes%>%
  filter(city %in% c(bikes%>%
                       count(city)%>%
                       filter(n > 90)%>%
                       arrange(n)%>%
                       pull(city)))
```


```{r echo =F}
bikes <- bikes%>%
  select(-crashyear)
```

### MODELING 

#### Training/Test Split 

```{r echo=F, message= F}
library(tidymodels)
library(rpart)
```

```{r}
bikes$ambulance <- as.factor(bikes$ambulance)
#SPLIT
set.seed(666)
# Create the balanced data split
bikes_split <- initial_split(bikes, 
                             prop = .75,
                             strata = ambulance)
# Extract the training and test set
bikes_train <- training(bikes_split)
bikes_test  <- testing(bikes_split)
```

#### Random Forest 
This is a go-to model for classification problems. Perfect balance between complexity and accuracy. 

#### Plot Variable Importance 
```{r}
# Specify a random forest
spec <- rand_forest() %>%
  set_mode("classification") %>%
  set_engine('ranger', importance = 'impurity')
# Train the forest
model <- spec %>%
  fit(ambulance ~.,
      data = bikes_train)
# Plot the variable importance
vip::vip(model, aesthetics =list(fill = '#31a7eb')) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
                        panel.background = element_blank(), axis.line = element_line(colour = "black"))
```

Feature importance helps us to identify features that are having the most impact on outcome variables. Even though it's a good indicator for the analysis, we have to be careful while interpreting these results. More domain knowledge is needed. 
Here we see that the time when a crash happened has the most effect on the severity of an accident. 

##### Evaluating accuracy 

```{r}
# Generate predictions
predictions <- predict(model,
                       new_data = bikes_test, type = 'class')
# Add the true outcomes
predictions_combined <- predictions %>% 
  mutate(true_class = bikes_test$ambulance)
# The confusion matrix
bikes_matrix <- conf_mat(data = predictions_combined,
                         estimate = .pred_class,
                         truth = true_class)
# Get accuracy of the model 
acc_auto <- accuracy(predictions_combined,
                     estimate = .pred_class, 
                     truth = true_class)
acc_auto$.estimate
```

Current accuracy is 0.7116345, which can be improved by tuning and using boosted-trees. But I would not be expecting a big increase in model performance since we are trying to predict real-life events involving humans and there's no machine learning algorithm that can predict human behaviour (yet).
